---
title: "Chart Toppers"
subtitle: "INFO 526 - Summer 2026 - Final Project"
author: 
  - name: "Gavin Caldwell"
    affiliations:
      - name: "School of Information, University of Arizona"
description: "This project dives into a data set detailing an extensive amount of information about every single song that has been ranked #1 on the Billboard Hot 100 list from 1958 to 2024"
format:
   html:
    code-tools: true
    code-overflow: wrap
    embed-resources: true
editor: visual
execute:
  warning: false
  echo: false
---

```{r}
#| label: package-setup
#| warning: FALSE
#| message: FALSE
#| echo: FALSE

# Set up packages and such
if (!require("pacman")) 
  install.packages("pacman")

# Install packages
pacman::p_load(tidyverse,
               dsbox,
               here,
               lubridate,
               stringr,
               patchwork)

# set wieght of code output
options(width = 65)

# set figure parameters
knitr::opts_chunk$set(
  fig.width = 7,
  fig.asp = 0.618,
  fig.retina = 3,
  fig.align = "center",
  dpi = 300
)

# read in csv for each question
billboard <- read.csv(here("data", "billboard.csv"))

```

## Abstract

This project aims to answer two questions regarding how the nature of American Popular music has changed over time. It does this via a data scientific approach, examining data on each #1 song as defined by the Billboard Hot 100 list from 1958 to 2024. The two questions answered by this project are "Has American popular music grown more homogeneous over time?" and "Has the diversity of record labels that produce American popular music decreased over time?".\
\
To answer question one, a chart showing the standard deviation of BPMs of #1 song over time, and a stacked area chart of Musical Keys of #1 songs were shown. To answer question 2,

## Introduction

The American Music industry is an incredibly influential source of culture and inspiration for millions if not billions of people around the globe, and it has changed a lot over the last 75 years. One can look to data science to help visualize and understand these changes. Using a dataset encompassing an impressive amount of information on each #1 song on the Billboard Hot 100 list from 1958 to 2024, this project aims to answer two questions about American Popular Music: "Has American popular music grown more homogeneous over time?" and "Has the diversity of record labels that produce American popular music decreased over time?".

The dataset used for this project is by and large the work of one person: Chris Dalla Riva. It includes information about each songs' musical structure, its genre, who performed it, who produced it, and even some obscure data such as if it was made for a movie or a play. This question of homogeneity in popular music will be answered by looking to each songs' musical structure as well as the organizations that produced it. The inspiration for these two questions come from the idea that if the number of distinct corporations that are actually producing popular music in America have decreased, it would lead to ultimately the most popular songs in America sounding the "same", and discouraging creativity and experimentation. The data presented here will aim to provide insight to this idea and whether it holds merit.

## Has American popular music grown more homogeneous over time?

### Introduction

### Approach

### Analysis

```{r}
#| label: bpm-viz
#| warning: FALSE
#| echo: TRUE

# read in data
bpm_length = billboard |> # data wrangling
  select("song", "date", "bpm", "length_sec") |> # isolating variables
  mutate(year = year(ymd_hms(date))) |> # convert to year 
  filter(year != 2025) |>
  pivot_longer( # pivoting longer for bpm and song length
    cols = c(bpm, length_sec),
    names_to = "category",
    values_to = "value"
  ) |>
  group_by(year, category) |> # grouping by year and category for Standard Deviation category
  summarise(sd = sd(value, na.rm = TRUE))

ggplot(bpm_length, aes(x = year, y = sd, color = category)) +
  geom_line() +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ category, ncol = 1) + 
  theme_minimal()

```

```{r}
#| label: keys-viz
#| warning: FALSE
#| echo: TRUE

keys_data <- billboard |> # data wrangling
  select("song", "date", "keys", "simplified_key") |> # isolating variables
  mutate(year = year(ymd_hms(date))) |> # convert to year 
  filter(year != 2025)

# count the use of keys
keys_count <- keys_data |>
  mutate(year_group = (year %/% 5) * 5) |>
  mutate(
    keys = str_remove_all(keys, "&"),
    keys = str_remove_all(keys, "%")) |> # cleaning up keys variable
  separate_rows(keys, sep = ";") |> # enumerate all keys used in song
  count(year_group, keys) |> # count the times a key was used in a given year
  complete(year_group, keys, fill = list(n = 0)) # fill in missing keys with 0 for plotting purposes

# count the use of multiple keys
multiple_keys_count <- keys_data |>
  group_by(year) |>
  summarize(multiple_keys = sum(
    simplified_key == "Multiple Keys", 
    na.rm = TRUE)
  )

ggplot(keys_count, aes(x = year_group, y = keys, fill = n)) +
  geom_tile() +
  scale_fill_viridis_c(option = "cividis") +
  theme_minimal()


ggplot(multiple_keys_count, aes(x =  year, y = multiple_keys)) +
  geom_line() +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal()
```

```{r}
#| label: keys-viz-part2
#| warning: FALSE
#| echo: TRUE

modern_keys_data <- billboard |> # data wrangling
  select("song", "date", "keys", "simplified_key") |> # isolating variables
  mutate(year = year(ymd_hms(date))) |> # convert to year 
  filter(year > 1990) |>
  mutate(year_group = (year %/% 5) * 5) |>
  mutate(
    keys = str_remove_all(keys, "&"),
    keys = str_remove_all(keys, "%")) |> # cleaning up keys variable
  separate_rows(keys, sep = ";") |> # enumerate all keys used in song
  add_count(keys, name = "count") |> # getting count of keys
  select(-count) |> # removing total count variable
  count(year_group, keys) |> # count the times a key was used in a given year
  complete(year_group, keys, fill = list(n = 0)) # fill in missing keys with 0 for plotting purposes
  

ggplot(modern_keys_data, aes(x = year_group, y = keys, fill = n)) +
  geom_tile() +
  scale_fill_viridis_c(option = "cividis") +
  theme_minimal()
```

### Discussion

## Has the diversity of record labels that produce American popular music decreased over time?

### Introduction

### Approach

### Analysis

```{r}
#| warning: FALSE
#| echo: TRUE

label_data <- billboard |> # data wrangling
  select("weeks_at_number_one", "date", "label", "parent_label") |> # isolating variables
  mutate(year = year(ymd_hms(date))) |> # convert to year 
  filter(year != 2025) |>
  separate_rows(label, sep = ";") 


# Find labels with fewer than 25 total weeks across all years
small_labels <- label_data |>
  group_by(label) |>
  summarise(
    total_weeks = sum(weeks_at_number_one),
    .groups = "drop"
  ) |>
  filter(total_weeks < 50) |>
  pull(label)

# filter out small labels and then proportions plot across all years
record_label <- label_data |>
  mutate(label = ifelse(label %in% small_labels, "Other", label)) |>
  group_by(year, label) |>
  summarise(
    weeks_at_number_one = sum(weeks_at_number_one),
    .groups = "drop"
  ) 

ggplot(record_label, aes(x = year, y = weeks_at_number_one, fill = label)) +
  geom_col(position = "fill") +
  scale_fill_viridis_d(option = "A") +
  theme_minimal()
  


```

### Discussion

## Conclusion
