---
title: "How the American Music Industry has Changed"
subtitle: "A look at #1 Hit Songs in the U.S."
author: 
  - name: "Gavin Caldwell"
    affiliations:
      - name: "School of Information, University of Arizona"
description: "This project dives into a data set detailing an extensive amount of information about every single song that has been ranked #1 on the Billboard Hot 100 list from 1958 to 2024"
format:
  html:
    code-tools: true
    code-overflow: wrap
    code-line-numbers: true
    embed-resources: true
editor: visual
code-annotations: hover
execute:
  warning: false
---

```{r}
#| label: load-pkgs
#| include: false
if (!require(pacman))
  install.packages("pacman")

# only download packages if needed, otherwise, load
pacman::p_load(
  tidyverse,
  here
)
```

## Dataset

The origins of this dataset are by and large the work of one person: Chris Dalla Riva. Here he is describing the origins of this impressive volume of information:

> 7 years ago, I decided that I was going to listen to every number one hit. Along the way, I tracked an absurd amount of information about each song. Using that information, I wrote a data-driven history of popular music covering 1958 through today.

His website, *Can't Get Much Higher*, where you can buy the data-science book he wrote based off his endeavor, and check out his other work, is linked [here](https://www.cantgetmuchhigher.com). The dataset was then uploaded to the TidyTuesday 2025 dataset by [Jen Richmond](https://github.com/jenrichmond).

This dataset interested me because I absolutely love music and am often fascinated by the developments of genres and music trends over time. Back in my undergraduate degree I took a gen-ed course entitled [*MUS109 - American Popular Music and Society*](https://catalog.arizona.edu/courses/0230671/general-aoYks), which dove into the most popular genre in American music of each decade starting in the early 1900's to today, and this dataset sparked a similar curiosity as that class did.

Here is a look at the first few rows of the data. A more detailed description can be found in the [data folder](https://github.com/INFO-526-SU26/final-project-gmcal213/blob/main/data/README.md) of this project's repository. There are are lot of different rows of individual songs of course, but there is also a substantial amount of columns representing all kinds of different information about each song.

```{r}
#| label: load-dataset
#| message: false

hot_100 <- read_csv(here("data", "billboard.csv"))

glimpse(hot_100)
```

## Questions

With this dataset I aim to answer the following two questions about the American music industry:

1.  Has American popular music grown more homogeneous over time?
2.  Has the diversity of record labels that produce American popular music decreased overtime?

## Analysis plan

The analysis for both of these questions will involve time along the x-axis, grouped by year. Drawing back on my previous experiences when studying this topic, I find it most interesting to observe trends as they occur at different points in history. This is the only variable that these two questions will share. There is a variable named "date" that informs when a song first hit #1 on the charts, and is captured in a standard "date" format as day-month-year (i.e. 03-Sep-76).

### Homogeneity in Popular Music

How do you measure how "similar" two songs are to each other? To do this, I will employ 3 variables that are labeled as "Musical Info" by the dataset and produce 2-3 charts. These variables are:

- Keys
- Simplified Key
- BPM

I considered using the "Lyrical Topic" variable as well, but it seems too subjective and not concrete, while the Musical Info variables are objective and lend itself much better to quantitative analysis.

The first chart will be taking a look at the BPM (beats per minute) of #1 songs over time. BPM is a measure of how fast or slow a song is. To measure how similar or dissimilar songs are growing over time with this metric I will plot the standard deviation of BPMs used over time.

Using the "Keys" and "Simplified Key" variables, I aim to show how the distribution of keys used over time has changed, and whether the distribution has grown wider or smaller. A musical key, is, very simply, a letter A-G (with sharp's and flat's) corresponding to one of the twelve notes of Western Music that determines a set of a chords and sounds a song should use to give the song a sense of musical cohesion. \
\
To do this, I will use a stacked area charts of keys used in a given years, faceting multiple charts at 5 year intervals to display the change over time. The question of Keys used is made a bit more interesting when one considers that certain songs will use multiple Keys, differentiated in the dataset as true "Multiple Keys" or if a change is just an "energy key change". My initial plan is when a song has multiple keys, the first key listed will be chosen, assuming that it is the primary key of the song. I will also use songs that are listed as having "Multiple Keys" as an auxiliary chart to enhance the main one by the plotting number of songs that use multiple keys over time.

### Diversity in Producing Popular Music

To measure the diversity of who is producing popular music in America, the following three variables will be used:

- Weeks at Number One
- Label
- Parent Label

Rather than simply counting how many times a Record Label company shows up when a new song hits #1, I want to use the "Weeks at Number One" variable to get a true sense of how much "market share" companies have over the time they have spent with one of their artists having the #1 song. The "Label" variable represents the primary record label that released the song, while "Parent Label" is a larger parent company that owns the record label, if any.

I will produce two charts to answer this question. Both charts will be a proportion bar chart chart of the weeks at number one held by either a record label or parent label over time. To avoid the chart being unnecessarily busy, I will likely need to filter out the one-off record labels, or create a new variable called "One-Time Labels" which would encompass all of the labels that made a one-time appearance at the #1 spot that year, and did not show up again. These labels would be filtered out in the final chart.

## Weekly Time Table

| Task Name | Status | Due | Priority | Summary |
|---------------|---------------|---------------|---------------|---------------|
| Determine Dataset for Project | **Complete in Week 1** | Week 1 | Low | Just getting familiar with the class and course materials here, found a dataset that I found intriuging. |
| Establish Project Repo + Start Proposal | **Complete Week 2** | Week 2 | High | Project repo cloned and cleaned up. Majority of proposal complete. |
| Finish Proposal + Peer Review | **In Progress -\> Will Need to Finish in Week 4** | Week 3 | Medium | Homework 1 took up more time than I anticipated. Will need a better plan going forward how to manage the Project and homework at the same time. |
| Address Peer Review + Start Question 1 | **Not Started** | Week 4 | Low |  |
| Finish Question 1 + Start Question 2 | **Not Started** | Week 5 | Medium |  |
| Finish Question 2 + Begin Write Up | **Not Started** | Week 6 | High |  |
| Finish Proposal/Peer Review | **Not Started** | Week 7 | High |  |
