---
title: "Chart Toppers - Proposal"
subtitle: "A look at #1 Hit Songs in the U.S."
author: 
  - name: "Gavin Caldwell"
    affiliations:
      - name: "School of Information, University of Arizona"
description: "This project dives into a data set detailing an extensive amount of information about every single song that has been ranked #1 on the Billboard Hot 100 list from 1958 to 2024"
format:
  html:
    code-tools: true
    code-overflow: wrap
    code-line-numbers: true
    embed-resources: true
editor: visual
code-annotations: hover
execute:
  warning: false
---

```{r}
#| label: load-pkgs
#| include: false
if (!require(pacman))
  install.packages("pacman")

# only download packages if needed, otherwise, load
pacman::p_load(
  tidyverse,
  here
)

# load dataset
hot_100 <- read_csv(here("data", "billboard.csv"))
```

## Dataset

The origins of this dataset are by and large the work of one person: Chris Dalla Riva. Here he is describing the origins of this impressive volume of information:

> 7 years ago, I decided that I was going to listen to every number one hit. Along the way, I tracked an absurd amount of information about each song. Using that information, I wrote a data-driven history of popular music covering 1958 through today.

His website, *Can't Get Much Higher*, where you can buy the data-science book he wrote based off his endeavor, and check out his other work, is linked [here](https://www.cantgetmuchhigher.com). The dataset was then uploaded to the [TidyTuesday 2025 Project](https://github.com/rfordatascience/tidytuesday/blob/main/data/2025/2025-08-26/readme.md) by [Jen Richmond](https://github.com/jenrichmond).

This dataset interested me because I absolutely love music and am often fascinated by the developments of genres and music trends over time. Back in my undergraduate degree I took a gen-ed course entitled [*MUS109 - American Popular Music and Society*](https://catalog.arizona.edu/courses/0230671/general-aoYks), which dove into the most popular genre in American music of each decade starting in the early 1900's to today, and this dataset sparked a similar curiosity as that class did.

Here is a look at the first few rows of the data. A more detailed description can be found in the [data folder](https://github.com/INFO-526-SU26/final-site-gmcal213/tree/main/data) of this project's repository. There are are ultimately `r nrow(hot_100)` different rows of individual songs, and `r ncol(hot_100)` columns representing all kinds of different information about each song.

```{r}
#| label: load-dataset
#| message: false

glimpse(hot_100)

```

## Questions

With this dataset I aim to answer the following two questions about the American music industry:

1.  Has the diversity of record labels that produce American popular music decreased over time?
2.  Has American popular music grown more homogeneous over time?

## Analysis plan

The analysis for both of these questions will involve time along the x-axis, grouped by year. Drawing back on my previous experiences when studying this topic, I find it most interesting to observe trends as they occur at different points in history. This is the only variable that these two questions will share. There is a variable named "date" that informs when a song first hits #1 on the charts, and is captured in a standard "date" format as day-month-year (i.e. 03-Sep-76).

### Diversity in Producing Popular Music

To measure the diversity of who is producing popular music in America, the following three variables will be used to produce 2 charts:

- Weeks at Number One
- Label
- Parent Label

Rather than simply counting how many times a Record Label company shows up when a new song hits #1, I will use the "Weeks at Number One" variable to get a true sense of how much "market share" companies have over the time they have spent with one of their artists at the #1 spot. The "Label" variable represents the primary record label that released the song, while "Parent Label" is a larger parent company that owns the record label, if any.

I will produce two charts to answer this question. Both charts will be a 100% proportion bar chart of the weeks at number one held by a record label and a parent label over time, grouped by 5 year intervals. To avoid the chart being unnecessarily busy, and highlight the centralization of record/parent labels, I will group the top majority of labels together and plot their share of the time at #1 against all of the other "Independent" labels.

### Homogeneity in Popular Music

How do you measure how "similar" two songs are to each other? To do this, I will employ 4 variables that are labeled as "Musical Info" by the dataset and produce 3 charts. These variables are:

- Keys
- Song Length in Seconds
- Simplified Key
- BPM

The first chart will be taking a look at the BPM (beats per minute) and song length in seconds of #1 songs over time. BPM is a measure of how fast or slow a song is, and song length is a measure how short or long a song is. To measure how similar or dissimilar #1 songs are growing over time with these metrics, I will plot the standard deviation of BPMs and song lengths over time on a line graph.

Using the "Keys" and "Simplified Key" variables, I aim to show how the distribution of keys used in #1 songs over time has changed, and whether the distribution has grown wider or smaller. A musical key, is, very simply, one of 24 possible symbols corresponding to the twelve notes of Western Music that determines a set of a chords and sounds a song should use.

To do this, I will use a heat map of the number of times each musical key was used in a #1 song, grouped by 5 year intervals. This will show if there are hot spots of keys favored by artists and songwriters. This question is made a bit more interesting when one considers that certain songs may use multiple keys, differentiated in the dataset as "Multiple Keys" from the simplified key variable. Further, a song can be labelled with what's known as just an "energy key change", in which case it is not counted as truly using multiple keys. If a song uses multiple keys, all instances of those keys will be accounted for in the heat map. I will use songs that are listed as having "Multiple Keys" on an auxiliary line graph by plotting the number of #1 songs that employ multiple keys over time.

## Weekly Time Table

| Task Name | Status | Due | Priority | Summary |
|---------------|---------------|---------------|---------------|---------------|
| Determine Dataset for Project | **Completed** | Week 1 | Low | Just getting familiar with the class and course materials here, found a dataset that I found intriuging. |
| Establish Project Repo + Start Proposal | **Completed** | Week 2 | High | Project repo cloned and cleaned up. Majority of proposal complete. |
| Finish Proposal + Peer Review | **Completed** | Week 3 | Medium | Homework 1 took up more time than I anticipated. Will need a better plan going forward how to manage the Project and homework at the same time. |
| Address Peer Review + Start Question 1 | **Peer review Completed, Write Up delayed** | Week 4 | Low | Addressed peer review comments, issues closed. |
| Finish Question 1 + Start Question 2 | **Proposal Completed, Question 1 Started** | Week 5 | Medium | Still struggling to keep up with homework, proposal is complete but will have to really focus on the write-up next week. |
| Finish Question 2 + Begin Presentation | **Write Up Complete, Presentation Started** | Week 6 | High | Finally got ahead of the homework, finished up write up and began presentation |
| Finish Proposal/Peer Review | **Not Started** | Week 7 | High |  |
